---
title: "Generalized additive models"
author: "A Irwin"
date: "2023-06-14"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, quietly = TRUE)
library(mgcv)
library(gratia)
```


Read data files.

```{r message=FALSE}
niskin <- read_csv("Cariaco/niskin-ss.csv")
diatom <- read_csv("Cariaco/diatoms-selected.csv") |>
  pivot_longer(d_1m:d_100m, names_to = "depth", values_to = "abundance") |>
  mutate(depth = str_remove(depth, "d_") |> str_remove("m") |> as.numeric(),
         present = abundance > 0)
diatom_environment <- diatom |> left_join(niskin, by = c("Cruise_ID_2" = "Cruise_ID", "depth")) |>
  mutate(log_abundance = log(abundance))
diatom_sum <- read_csv("Cariaco/diatoms-selected-sum.csv") |>
  pivot_longer(d_1m:d_100m, names_to = "depth", values_to = "abundance") |>
  mutate(depth = str_remove(depth, "d_") |> str_remove("m") |> as.numeric(),
         high_abundance = abundance > 0.1)
diatom_sum_environment <- diatom_sum |> left_join(niskin, by = c("Cruise_ID_2" = "Cruise_ID", "depth")) |>
  mutate(log_abundance = log(abundance))
```

Make a generalized linear model to predict presence or absence of a species.

```{r}
model1 <- gam(present ~ s(NO3_NO2_USF) + s(PO4_USF) + s(SiO4_USF) + 
                s(Temperature) + s(Salinity_bottles) + 
                s(pH) + s(Month_local, bs = "cc"), 
             data = diatom_environment |> 
               # filter(ScientificName_accepted == "Dactyliosolen fragilissimus"),
               filter(ScientificName_accepted == "Leptocylindrus danicus"),
               # filter(ScientificName_accepted == "Navicula sp."),
             family = "binomial",
             select = TRUE
             )
summary(model1)
```

Draw response functions.

```{r}
draw(model1, select = c(1:2, 4:5, 7))
```

```{r}
draw(model1, select = c(1:2, 4:5, 7), residuals = TRUE)
```


## Model total diatom abundance

```{r}
diatom_sum_environment_ss <- diatom_sum_environment |>
  filter(abundance > 0)
model3 <- gam(log_abundance ~ s(NO3_NO2_USF) + s(PO4_USF) + s(SiO4_USF) +
                s(Temperature) + s(Salinity_bottles) + s(pH) + 
                s(Month_local, bs = "cc") + depth + s(Year_local), 
             data = diatom_sum_environment_ss,
             family = "gaussian",
             select = TRUE
             )
summary(model3)
```

```{r}
draw(model3, select = c(3:4, 6:8))
```


```{r}
draw(model3, select = c(3:4, 6:8), residuals = TRUE)
```

## Utility functions

Show the basis functions for a predictor from the previous model.


```{r}
basis(s(Temperature), data = diatom_sum_environment_ss |> filter(!is.na(Temperature))) |>
  ggplot(aes(y = value, x = Temperature, group = bf, color = bf)) + 
  geom_line() + 
  theme_bw()
```


This is piecewise linear because it is only evaluated at the integers 1 to 12. Use a different variable, or add `bs="cc"` to the temperature example above to see smooth versions of the basis functions.

```{r}
basis(s(Month_local, bs = "cc"), data = diatom_sum_environment_ss) |>
  ggplot(aes(y = value, x = Month_local, group = bf, color = bf)) + 
  geom_line() + 
  theme_bw()
```
