---
title: "Linear models"
author: "A Irwin"
date: "2023-06-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse, quietly = TRUE)
library(broom)
```

Read data files.

```{r message=FALSE}
niskin <- read_csv("Cariaco/niskin-ss.csv")
diatom_sum <- read_csv("Cariaco/diatoms-selected-sum.csv") |>
  pivot_longer(d_1m:d_100m, names_to = "depth", values_to = "abundance") |>
  mutate(depth = str_remove(depth, "d_") |> str_remove("m") |> as.numeric() )
diatom_sum_environment <- diatom_sum |> left_join(niskin, by = c("Cruise_ID_2" = "Cruise_ID", "depth")) |>
  mutate(log_abundance = log(abundance))
```

Make a linear model for total abundance.

```{r}
model1 <- lm(abundance ~ NO3_NO2_USF + PO4_USF + SiO4_USF + Temperature + Salinity_bottles + pH, 
             data = diatom_sum_environment |> filter(depth == 1))
summary(model1)
```

We get a significant effect of temperature and pH. About 21% of variation in the data are explained by these variables.

Try log abundance.

```{r}
diatom_sum_environment_1m_no_zero <- diatom_sum_environment |> filter(depth == 1, abundance > 0)
model2 <- lm(log_abundance ~ NO3_NO2_USF + PO4_USF + SiO4_USF + Temperature + Salinity_bottles + pH, 
             data = diatom_sum_environment_1m_no_zero)
summary(model2)

```

Same variables are significant, but we explain 49% of the total variance. The coefficients are different because we are predicting log abundance instead of abundance, but the signs are the same.

Use `broom` pacakge functions to extract coefficients and model diagnositics into a data frame.

```{r}
tidy(model2, conf.int = TRUE)
glance(model2)
```

## Plots

Show predicted vs. observed log diatom abundance.

```{r}
augment(model2) |>
  ggplot(aes(log_abundance, .fitted)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  theme_bw()
```

Show residuals as a function of fitted values; should be a 'null plot'.

```{r}
augment(model2) |> # change to model1 to see why predicting abundance is a bad idea
  ggplot(aes(.fitted, .resid)) + # looks great
  geom_point() +
  geom_abline(intercept = 0, slope = 0) + 
  theme_bw()
```

Show the quantile-quantile plot of the residuals to visually test if they are Normal.

```{r}
augment(model2) |>
  ggplot(aes(sample = .resid)) + # looks great
  stat_qq() + stat_qq_line() + 
  theme_bw()
```
Is there a relationship between observed values and residuals? Sadly, yes.

```{r}
augment(model2) |>
  ggplot(aes(log_abundance, .resid)) + # clear bias in residuals
  geom_point() + 
  theme_bw()
```


## Exercises

Repeat this analysis for another group (dinoflagellates, coccolithophores).

Repeat this analysis for a species observed frequently (or rarely if you want to see a terrible model.)

