---
title: "cariaco"
author: "AJ Irwin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(naniar)
library("worrms")
```

## Phytoplankton

```{r message=FALSE}
fn <- "phytoplankton.txt"
h <- readLines(fn, n = 1) # read header
col_names_v <- str_squish(h) |> str_split(" ")
phyto <- read_fwf(fn, na = "nd", skip = 1,
                  fwf_empty(file = fn, col_names = col_names_v[[1]]))
 # get a warning about a problem in column 19 of row 3314; ignore for now
```

Some exploratory plots.

```{r}
# gg_miss_case(phyto)
phyto |> select(-(Cruise:ScientificName_accepted)) |> gg_miss_var(show_pct = TRUE)
frac <- function(x) x - floor(x)
phyto |> count(ScientificName_accepted)
ss1 <- phyto |> 
  filter(Year_local > 1995, Year_local < 2016) |>
  mutate(dd = decimal_date(Datetime_local)) |>
  group_by(Cruise, dd, Year_local) |>
  summarize(across(starts_with("d_"), sum), .groups = "drop") |>
  pivot_longer(starts_with("d_"), names_to = "depth") |>
  mutate(depth = str_remove_all(depth, "(d_|m)") |> as.numeric()) |> 
  filter(value > 0) 
ss1 |>
  ggplot(aes(dd, value, color = depth)) + geom_point() +
  # geom_smooth(formula = y~ s(x), method = "gam", aes(group = depth), se = FALSE) +
  geom_smooth(formula = y~ x, method = "loess", span = 0.2, aes(group = depth), se = FALSE) +
  scale_y_log10() + 
  theme_bw() +
  labs(y = "Phytoplankton total abundance (#/L)",
       x = "Sample date", 
       color = "Depth (m)")
ss1 |>
  ggplot(aes(frac(dd), value, color = depth)) + geom_point() +
  # geom_smooth(formula = y~ s(x), method = "gam", aes(group = depth), se = FALSE) +
  # geom_smooth(formula = y~ x, method = "loess", span = 0.2, aes(group = depth), se = FALSE) +
  scale_y_log10() + 
  facet_wrap(~ Year_local) +
  theme_bw() +
  labs(y = "Phytoplankton total abundance (#/L)",
       x = "Fractional year", 
       color = "Depth (m)")
ss1 |>
  filter(depth < 2 ) |>
  ggplot(aes(frac(dd), value, color = Year_local)) + geom_point() +
  geom_smooth(formula = y~ s(x), method = "gam", se = TRUE) +
  # geom_smooth(formula = y~ x, method = "loess", span = 0.2, se = FALSE) +
  scale_y_log10() + 
  theme_bw() +
  labs(y = "Phytoplankton total abundance (#/L)",
       x = "Fractional year", 
       color = "Year")
```

Get the taxonomic hierarchy by looking up AphiaID at marinespecies.org. I'll do this once and store the result so you don't need to repeat it. Add the class part of the taxonomic hierarchy so we can aggregate counts.

```{r}
## taxonomic_information <- wm_recordtax(id = unique(phyto$AphiaID))
## write_csv(bind_rows(taxonomic_information), "taxonomic-information.csv")
taxonomic_information <- read_csv("taxonomic-information.csv")
phyto_class <- phyto |> left_join(taxonomic_information |> select(AphiaID, class), by = "AphiaID")
phyto_class_aggregate <- phyto_class |> 
  select(-Latitude, -Longitude, -AphiaID, -Leg, -Cast, -Day_local) |>
  group_by(Cruise, Cruise_ID_2, Month_local, Year_local, class) |>
  summarize(across(where(is.numeric), sum), .groups = "drop")
phyto_ss <- phyto_class_aggregate |>
  filter(class %in% c("Bacillariophyceae", "Coccolithophyceae", "Dinophyceae"))
```

Plot some time series and histograms.

```{r}
phyto_ss |> ggplot(aes(Cruise, d_1m)) +
  geom_point() +
  facet_wrap(~class) +
  scale_y_log10() + 
  theme_bw()
phyto_ss |> ggplot(aes(d_1m)) +
  geom_histogram() +
  facet_wrap(~class) +
  scale_x_log10() + 
  theme_bw()

```

Write these counts to file for later analysis.

```{r}
write_csv(phyto_ss, "phytoplankton-class-aggregate.csv")
```

## Zooplankton

```{r}
# zoo <- read_table("zooplankton.txt", na = "nd")
fn <- "zooplankton.txt"
h <- readLines(fn, n = 1)
col_names_v <- str_squish(h) |> str_split(" ")
zoo <- read_fwf(fn, na = "nd", skip = 1,
                fwf_empty(file = fn, col_names = col_names_v[[1]]))
```

Exploratory plots.

```{r}
gg_miss_case(zoo)
zoo |> select(-(Cruise:Mesh_Size)) |> gg_miss_var(show_pct = TRUE)
zoo |>
  select(Date, Month, Mesh_Size, BIOMASS, ASH, COPEPODS, CHAETOGNATHA, OSTRACODA, FORAMINIFERA, AMPHIPODS, EUPHAUSIIDS, MEDUSA, PTEROPODS, SALPS, CTENOPHORES) |>
  # pivot_longer(COPEPODS:JUV_GASTROPOD) |>
  pivot_longer(BIOMASS:CTENOPHORES) |>
  ggplot(aes(Month, value, color = factor(Mesh_Size))) + 
  # geom_point() +
  stat_summary(fun.data = "mean_cl_boot", size = 0.5) +  # median_hilow, mean_cl_boot
  facet_wrap(~name, scales = "free_y") +
  scale_y_log10() + 
  theme_bw() +
  labs(color = "Mesh Size", 
       y = "Abundance or Biomass (# or mg / m3)")
```

Time series

```{r}
zoo |>
  select(Date, Month, Mesh_Size, BIOMASS,  COPEPODS, CHAETOGNATHA, OSTRACODA, FORAMINIFERA, AMPHIPODS,  PTEROPODS, SALPS) |>
  # pivot_longer(COPEPODS:JUV_GASTROPOD) |>
  pivot_longer(BIOMASS:SALPS) |>
  ggplot(aes(Date, value, color = factor(Mesh_Size))) + 
  geom_point(size = 0.4) +
   # stat_summary(fun.data = "mean_cl_boot", size = 0.2) +  # median_hilow, mean_cl_boot
  facet_wrap(~name, scales = "free_y") +
  scale_y_log10() + 
  theme_bw() +
  labs(color = "Mesh Size", 
       y = "Abundance or Biomass (# or mg / m3)") + 
  theme(legend.position = c(0.8, 0.15))
ggsave("cariaco-zoo-time.pdf", height = 4.5, width = 7)
```

Heatmap

```{r}
zoo |>
  pivot_longer(COPEPODS:JUV_GASTROPOD) |>
  filter(Mesh_Size == 200) |>
  ggplot(aes(Cruise, name, fill = log10(value))) + 
  geom_tile() + 
  theme_bw() +
  labs(y="")
```

## Niskin bottle data

```{r}
fn <- "niskin.txt"
h <- readLines(fn, n = 1)
col_names_v <- str_squish(h) |> str_split(" ")
niskin <- read_fwf(fn, na = "nd", skip = 1,   # can't use a compressed file here
                   fwf_empty(file = fn, col_names = col_names_v[[1]]))
  # ignore warning about 14 parsing errors in rows 4, 6, 13, 14, 17, 44.
niskin |> select(-(Cruise_number:Depth_real)) |> gg_miss_var(show_pct = TRUE)
```

Some exploratory plots.

```{r}
niskin |> filter(Depth_target == 1) |> 
  select(Year, Month, NO3_NO2_USF, PO4_USF, SiO4_USF, Temperature, Salinity_bottles, 
         POC_ug_kg, PON_ug_kg, Chlorophyll) |>
  pivot_longer(NO3_NO2_USF:Chlorophyll) |>
  ggplot(aes(x = factor(Month), # Year + (Month-0.5)/12,
             y = value)) +
  # geom_point() +
  geom_boxplot() + 
  facet_wrap(~ name, scales = "free_y") +
  theme_bw()

```


## Correlations in environment

```{r warning=FALSE}
library(GGally)
ggpairs(niskin |> select(NO3_NO2_USF, PO4_USF, SiO4_USF, Temperature, Salinity_bottles, pH, TCO2, Depth_real) |>
          filter(Depth_real <= 100),
        progress = FALSE)

```

## Organize subset of data for course

Select a few species and a few environmental variables for use in Lesson 1.

Exercise: Add functional group annotations by looking up taxa in MarineSpecies.org.

```{r}
select_taxa <- c("Bacillariophyceae", "Dinoflagellata", 
                 "Thalassiosira eccentrica", "Thalassiosira gravida",
                 "Gyrodinium pingue", "Prorocentrum micans")
phyto_ss <- phyto |> rename(Year = Year_local, Month = Month_local, Day = Day_local,
                Cruise_ID = Cruise_ID_2) |>
  select(Cruise, Cruise_ID, Year, Month, Datetime_local, ScientificName_accepted,
         d_1m, d_7m) |>
  # count(ScientificName_accepted) |> arrange(-n)
  filter(ScientificName_accepted %in% select_taxa) |>
  group_by(Cruise, Cruise_ID, Year, Month, Datetime_local, ScientificName_accepted) |>
  summarise(d_1m = sum(d_1m),
            d_7m = sum(d_7m),
            .groups = "drop") |>
  pivot_longer(d_1m:d_7m, names_to = "depth", values_to = "abundance") |>
  mutate(depth = str_extract(depth, "[0-9]") |> as.numeric() )
zoo_ss <- zoo |> 
  select(Cruise, Cruise_ID, Date, Mesh_Size, TOTAL_DENSITY, BIOMASS, COPEPODS, CHAETOGNATHA)
niskin_ss <- niskin |> rename(Cruise_ID = Cruise_ID_2, depth = Depth_target) |>
  filter(depth < 10) |>
  select(Cruise_ID, depth, 
         NH4_USF, NO2_USF, NO3_NO2_USF, PO4_USF, SiO4_USF, pH, Alkalinity_umol_kg,
         TCO2, Salinity_bottles, Temperature, POC_ug_kg, PON_ug_kg, TPP, DOC, PrimaryProductivity,
         Chlorophyll)
write_csv(phyto_ss, "phyto-ss.csv")
write_csv(zoo_ss, "zoo-ss.csv")
write_csv(niskin_ss, "niskin-ss.csv")
# full_join(niskin_ss, phyto_ss, by = c("Cruise_ID", "depth") )
# full_join(niskin_ss, zoo_ss, by = c("Cruise_ID") )
```


