---
title: "Species distribution models"
format: html
editor: source
---

```{r echo=FALSE, message=FALSE}
library(tidyverse, quietly = TRUE)
library(glmnet)
library(glmnetUtils)
library(mgcv)
library(broom)
library(gratia)
```
## Goals

* Learn ways to apply linear regression to the question: What determines the abundance of plankton species?


## Oceanographic context

Phytoplankton and small zooplankton have short generation times and respond quickly to environmental forcing. We usually observe higher phytoplankton biomass where there are more resources and growing conditions are good. Try to make a statistical model predicting the abundance of a plankton species assuming the population is in steady state with its environment.

Key objections: advection of populations will mean the cells did not necessarily grow where they were found; predator-prey dynamics mean that biomass of prey might be low even if conditions for growth are excellent, because conditions for growth of the predator may also be excellent.

What can we expect to learn: variables that strongly or weakly influence abundance or presence of a species or group of species; differences in these effects across taxa; characterization of the niche of a taxon.

## Cariaco time series

The CARIACO project (Carbon Retention in a Colored Ocean) ran for about 20 years from 1995-2017. Samples were collected in most months, with an increasing number of gaps near the end of the project. A primary goal of the project was to better understand carbon export in this part of the ocean which is notable for its well preserved and laminated sediments. Here we will use the samples of phytoplankton and zooplankton from the upper mixed layer.

Data are archived on BCO-DMO. Here we use a subset of the full dataset. I encourage you to look at the full dataset, for example, to access data from depths below 7 m.

```{r}
phyto <- read_csv(here::here("data/Cariaco", "phyto-ss.csv"))
niskin <- read_csv(here::here("data/Cariaco", "niskin-ss.csv"))
```

Average the top two depths. Select one species. Combine phytoplankton count data and environment data.

```{r}
one_spp <- full_join(
  phyto |> filter(ScientificName_accepted == "Thalassiosira gravida") |>
    group_by(Cruise_ID, Year, Month, ScientificName_accepted) |>
    summarize(abundance = mean(abundance, na.rm=TRUE), .groups = "drop"),
  niskin |> group_by(Cruise_ID) |>
    summarize(across(is.numeric, mean, na.rm=TRUE), .groups = "drop") |>
    select(-depth),
  by = "Cruise_ID"
) |>
  mutate(log_abundance = log(abundance),
         presence = abundance > 0)
```

Make a few exploratory plots to show how abundance varies with temperature, chlorophyll.

```{r}
one_spp |> ggplot(aes(Temperature, abundance)) +
  geom_point() +
  scale_y_log10() +
  theme_bw()
```

```{r}
one_spp |> ggplot(aes(Chlorophyll, abundance)) +
  geom_point() +
  scale_y_log10() +
  theme_bw()
```

```{r}
one_spp |> ggplot(aes(NO3_NO2_USF, abundance)) +
  geom_point() +
  scale_y_log10() +
  theme_bw()
```


```{r}
one_spp |> ggplot(aes(Month, abundance)) +
  geom_point() +
  scale_y_log10() +
  theme_bw()
```




## Linear models

What factors do we know affect phytoplankton abundance, that are available at this site? Physical conditions (temperature, salinity, light, pH) and resource availability (macronutrient concentrations.)

Start with a simple approach: look for a linear relationship between the physical and chemical predictors and the abundance of a species.

For comparison, try making a logistic regression model of the presence or absence of a species.

Problem: predictors are correlated. Show heatmap.

Remove a predictor (important or unimportant). Look to see how the coefficients change.

```{r}
model1 <- lm(abundance ~ Year + Month + NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Salinity_bottles + Temperature + Chlorophyll,
             data = one_spp)
summary(model1)
model2 <- lm(log_abundance ~ Year + Month + NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Salinity_bottles + Temperature + Chlorophyll,
             data = one_spp |> filter(abundance>0))
summary(model2)
```

Drop out some variables.

```{r}
model3 <- lm(log_abundance ~ Month + NO3_NO2_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Temperature + Chlorophyll,
             data = one_spp |> filter(abundance>0))
summary(model3)
```

### Presence-absence model

```{r}

model4 <- glm(presence ~ Month + NO3_NO2_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Temperature + Chlorophyll,
             data = one_spp,
             family = binomial)
summary(model4)
tidy(model4, conf.int = TRUE)
```


Predictions. `type.predict` can be `link` (meaning the prediction is "on the scale" of the predictors -- the logit of the predicted probability) or `response` (the prediction is on the scale of the response variable, a probability between 0 and 1.)

```{r}
augment(model4, newdata = one_spp, type.predict = "response") |> 
  ggplot(aes(.fitted, fill = presence)) +
  geom_histogram() + 
  theme_bw()
augment(model4, newdata = one_spp, type.predict = "link") |> 
  count(presence, .fitted > 0)
# augment(model4, newdata = one_spp, type.predict = "response") |> 
#   count(presence, .fitted > 0.5)
```

## Generalized linear models and LASSO/ElasticNet

Use variable selection with LASSO to build the same models, but with a penalty for variables that have little effect on model performance.

`glmnet` does not have a formula interface, but the package `glmnetUtils` provides one. With just `glmnet` you need to make a vector of the response data (abudance or log abundance) and a matrix of the predictors (nutrients, temperature, etc.) There is a concise [overview](https://cran.r-project.org/web/packages/glmnetUtils/vignettes/intro.html) of the basic functions of `glmnetUtils`.

```{r}
glm_model1 <- glmnet(abundance ~ Year + Month + NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Salinity_bottles + Temperature + Chlorophyll,
             data = one_spp, 
             alpha = 0.5)
glm_model1
glm_model2 <- glmnet(log_abundance ~ Year + Month + NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + TCO2 + Salinity_bottles + Temperature + Chlorophyll,
             data = one_spp |> filter(abundance>0))
glm_model2
```

It's important to have no missing data in the data used for predictions. If you do, you get no prediction and the rows of the data and the predictions won't match.

```{r}
one_spp_ss <- one_spp |> filter(!is.na(Year + Month + 
    NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + 
    TCO2 + Salinity_bottles + Temperature + Chlorophyll))
predictions1 <- predict(glm_model1, 
                        newdata = one_spp_ss)
bind_cols(one_spp_ss, predictions1) |>
  ggplot(aes(abundance, s50)) + 
  geom_point() + 
  scale_x_log10() + scale_y_log10() + 
  theme_bw()
```


```{r}
predictions2 <- predict(glm_model2, 
                        newdata = one_spp_ss)
bind_cols(one_spp_ss, predictions2) |>
  ggplot(aes(log_abundance, s50)) + 
  geom_point() + 
  geom_abline(aes(intercept = 0, slope = 1), color = "blue", linetype = "dotted") + 
  # scale_x_log10() + scale_y_log10() + 
  theme_bw()
```

### Presence-absence

```{r}
glm_model3 <- glmnet(presence ~ Year + Month + NH4_USF + NO3_NO2_USF + PO4_USF + SiO4_USF + pH + Alkalinity_umol_kg + 
                       TCO2 + Salinity_bottles + Temperature + Chlorophyll,
             data = one_spp_ss,
             family = binomial)
glm_model3
```
```{r}
predictions3 <- predict(glm_model3, 
                        newdata = one_spp_ss)
bind_cols(one_spp_ss, predictions3) |>
  ggplot(aes(fill = presence, x= s50)) + 
  geom_histogram() + 
  theme_bw()
bind_cols(one_spp_ss, predictions3) |> #  |> as_tibble() |> 
  count(presence, s50 > 0)
```

True skill statistic  (TSS) is number correct divided by total number. TSS takes into account both omission and commission errors, and success as a result of random guessing, and ranges from −1 to +1, where +1 indicates perfect agreement and values of zero or less indicate a performance no better than random. TSS is not affected by prevalence.

First, construct the confusion matrix with the following entries:

* `a` true positives
* `b` false positives (predicted presence, observed absence)
* `c` false negatives (predicted absence, observed presence)
* `d` true negatives

```{r}
score_f <- function(a,b,c,d) {
  n = a+b+c+d
  sensitivity = a/(a+c)
  specificity = d/(b+d)
  x = (a+b)*(a+c) + (c+d)*(d+b)
  list(overall_accuracy = (a+d)/n,
    sensitivity = sensitivity,
    specificity = specificity,
    kappa =  ( (a+d)/n - x/(n^2) ) / ( 1 - x/(n^2) ),
    TSS = sensitivity + specificity - 1)
}
score_f(9, 4, 15, 137)
```


TSS reference: https://besjournals.onlinelibrary.wiley.com/doi/10.1111/j.1365-2664.2006.01214.x

## Generalized additive models

What about non-linear responses?


```{r}
gam_model1 <- gam(log_abundance ~ s(Month, bs = "cc", k = 5)  + 
                    s(NO3_NO2_USF, k = 3) + s(SiO4_USF, k = 3) + 
                    # s(pH, k = 3) + s(Chlorophyll, k = 3) +
                    s(TCO2, k = 3) + s(Temperature, k = 3) ,
             data = one_spp |> filter(abundance>0) ,
             select = TRUE)
summary(gam_model1)
draw(gam_model1, residuals = TRUE)

gam_model1a <- gam(log_abundance ~ s(Month, bs = "cc", k = 5)  + 
                    s(NO3_NO2_USF, k = 3) + s(SiO4_USF, k = 3) + 
                    s(Temperature, k = 3) ,
             data = one_spp |> filter(abundance>0) ,
             select = TRUE)
summary(gam_model1a)
draw(gam_model1a, residuals = TRUE)
```



```{r}
gam_model3 <- gam(presence ~ s(Year) + s(Month, bs = "cc", k= 5)  + 
                    s(NH4_USF) + s(NO3_NO2_USF) + s(PO4_USF) + s(SiO4_USF) + s(pH) + s(Alkalinity_umol_kg) + 
                       s(TCO2) + s(Salinity_bottles) + s(Temperature) + s(Chlorophyll),
             data = one_spp_ss,
             select = TRUE, 
             family = binomial)
summary(gam_model3)

draw(gam_model3, select = c("s(NO3_NO2_USF)", "s(Temperature)", "s(Salinity_bottles)", "s(Chlorophyll)"  ))


# Dropping variables that were shrunk to zero has no effect
gam_model3a <- gam(presence ~  s(Month, bs = "cc", k= 5)  + 
                    s(NO3_NO2_USF) + s(PO4_USF) + s(SiO4_USF) +  
                       s(TCO2) + s(Salinity_bottles) + s(Temperature) + s(Chlorophyll),
             data = one_spp_ss,
             select = TRUE, 
             family = binomial)
summary(gam_model3a)

draw(gam_model3a, select = c("s(NO3_NO2_USF)", "s(Temperature)", "s(Salinity_bottles)", "s(Chlorophyll)"  ))

```




## Abundances
What to do with zero abundances?
Logistic regression
Functional groups
Log ratios

## Challenge exercise

* Make a two step model. Make two models: one for the probability of presence and one for the log abundance, when abundance is non-zero. Then make a prediction for all locations. Use the predicted log abundance times the probability of presence.

## Extensions

* Phylogeny, abundance of other taxa, time-dependent dynamics
* Non-linear models with interactions: random forests, XGBoost, neural networks


